---
title: 'Time Series: Brooklyn'
author: "Dirk Hartog"
date: "2025-10-29"
output: html_document
---

```{r load libraries}
library(tidyverse)
library(fpp3)
library(fable)
library(seasonal)
library(GGally)
```

# INTRODUCTION

## Load data 

```{r}
# Read data in from github
daily_counts_type <- read_csv("https://raw.githubusercontent.com/D-hartog/DATA698/refs/heads/main/Project%20files/Data/brooklyn_dailyCIP_counts.csv")
head(daily_counts_type)
```

```{r}
daily_counts_type %>% group_by(nypd_pct_cd, incident_date) %>% 
  summarize(num_cip_types = n()) %>% filter(num_cip_types <=1)
```


```{r cip counts, fig.width = 10, fig.height = 8}
daily_counts_type %>% 
  mutate(call_type = case_when(
    cip_jobs == "Non CIP" ~ 1,
    cip_jobs == "Non Critical" ~ 2,
    cip_jobs == "Serious" ~ 3,
    cip_jobs == "Critical" ~ 4)) %>% 
  ggplot(aes(x = call_type, y = counts)) + geom_bar(stat = "identity") +
  facet_wrap(~ nypd_pct_cd, nrow = 5, ncol = 6, scales = "free") + labs(title = "Values") +  
  labs(title = "Precinct Call Count By Type",
       y = "", x = "Call Type") + 
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "lightgrey"),
        panel.grid.major = element_line(color = "grey", linewidth = 0.5))
  
```

```{r recode cip}
# Re-code call type into two distinct categories 
daily_counts_type <- daily_counts_type %>% 
  mutate(call_type = case_when(
    cip_jobs == "Non CIP" ~ "Non CIP",
    cip_jobs == "Non Critical" ~ "CIP",
    cip_jobs == "Serious" ~ "CIP",
    cip_jobs == "Critical" ~ "CIP"))

unique(daily_counts_type$call_type)

daily_counts_type_grouped <- daily_counts_type %>% group_by(nypd_pct_cd, incident_date, call_type) %>% summarise(count = sum(counts)) %>% ungroup()

# Use grouped data to see which precincts and dates contain only one type of call 
daily_counts_type_grouped %>% group_by(nypd_pct_cd, incident_date) %>% 
  summarize(num_cip_types = n()) %>% filter(num_cip_types <= 1)
```

## Feature Engineering

**Holiday Data**

```{r Read in holiday data}
holidays <- read_csv("https://raw.githubusercontent.com/D-hartog/DATA698/refs/heads/main/Project%20files/Data/Holidays_cleaned.csv")
head(holidays)
holiday_dates <- unique(holidays$Date)

```

### Weater Data 

```{r}
weather_data <- read_csv("https://raw.githubusercontent.com/D-hartog/DATA698/refs/heads/main/Project%20files/Data/Weather_data_cleaned.csv")

head(weather_data)
```

```{r recode weather}

weather_data <- weather_data %>% mutate(Prec = as.factor(ifelse(Precipitation <= 0.01, 0, 1)),
                                SF = as.factor(ifelse(Snowfall <= 0.1 , 0, 1)),
                                SD = as.factor(ifelse(SnowDepth <= 1, 0, 1)))

```


```{r create time series}
# Create tsibble
daily_counts_type_ts <- as_tsibble(daily_counts_type_grouped,  
                                   index = incident_date, 
                                   key = c(nypd_pct_cd, call_type))

# Aggregate keys 
daily_counts_type_ts <- daily_counts_type_ts %>% 
  aggregate_key(nypd_pct_cd * call_type, call_counts = sum(count))

# Engineer date features for possible use 
daily_counts_type_ts <- daily_counts_type_ts %>% mutate(month = factor(month(incident_date)),
                      dow = factor(weekdays(incident_date)),
                      weekend = factor(ifelse(wday(incident_date) %in% c(1,7), 1, 0)),
                      holiday = factor(ifelse(incident_date %in% holiday_dates, 1, 0))
)

daily_counts_type_ts <- daily_counts_type_ts %>% 
  left_join(weather_data, by = join_by(incident_date == Date))

daily_counts_type_ts
```



```{r}
# Check for missing values in the time series 

daily_counts_type_ts %>% count_gaps(.full = TRUE)
```

```{r }
# Visualize the gaps in precinct - 5
daily_counts_type_ts %>% count_gaps(.full = TRUE) %>% 
  mutate(call_type = factor(call_type), nypd_pct_cd = factor(nypd_pct_cd)) %>% 
  ggplot(aes(x = nypd_pct_cd, colour = call_type)) +
  geom_linerange(aes(ymin = .from, ymax = .to)) +
  geom_point(aes(y = .from)) +
  geom_point(aes(y = .to)) +
  coord_flip() +
  labs(title = "Time Gaps by Call Type", x = "Precinct", y = "") +
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "lightgrey"),
        panel.grid.major = element_line(color = "grey", linewidth = 0.5),
        legend.position = "bottom")
```

```{r fill gaps}
#daily_counts_type_ts <- daily_counts_type_ts %>% filter(nypd_pct_cd != 22)

daily_counts_type_ts <- daily_counts_type_ts %>% fill_gaps(.full = TRUE, call_counts = 1)
```


# HIERARCHICAL FORECASTING

## Forecasting total call counts by type

Fit three base models 
- Auto ETS
- Auto ARIMA 
- Auto ARIMA exo

Reconcile using MinT method

1. Fit with training data up to May 31st, 2025 - forecasting 30 days 
2. Fit with training data up to June 30th, 2024 - forecasting 365 days
**3. Fit with training data up to June 23rd, 2025 - forecasting for 7 days**


```{r fit hierarchical model}
fit_type <- daily_counts_type_ts %>% 
  filter(incident_date <= "2025-06-23") %>% 
  model(base_ets = ETS(call_counts),
        base_arima = ARIMA(call_counts,
                           approximation = TRUE),
        arima_weather = ARIMA(call_counts ~ dow + AvgTemperature + Prec + SF,
                           approximation = TRUE)) %>% 
  reconcile(
    mint_ets = min_trace(base_ets, method = "mint_shrink"),
    mint_arima = min_trace(base_arima, method = "mint_shrink"),
    mint_weather = min_trace(arima_weather, method = "mint_shrink")
  )
```


```{r predictor sig}
arima_models <- c(4,5,7)
fit_colnames <- colnames(fit_type)
precincts <- fit_type$nypd_pct_cd

precinct <- c()
mod_name <- c()
predictor_list <- c()

for(p in 1:nrow(fit_type)){
  for(m in arima_models){
    # get the term summary 
    model_terms <- fit_type[[m]][[p]][["fit"]][["par"]]
    # filter for terms that are non significant
    non_sig <- model_terms %>% filter(p.value > 0.05)
    # get predictor names
    term_names <- non_sig$term
    # print the precinct, model name and the terms that are not significant
    print(paste0("Precinct: ", precincts[p], " | ", "Model: ", fit_colnames[m], " | ", "Predictors: ", list(term_names)))
    precinct <- append(precinct, precincts[p])
    mod_name <- append(mod_name, fit_colnames[m])
    predictor_list <- append(predictor_list, list(term_names))
  }
}

tibble(Precinct = precinct,
       Model = mod_name,
       Predictors = predictor_list)
```

```{r Metrics}
glance(fit_type)
```

```{r save fit}
write_csv(fit_type, "brooklyn_fit_type.csv")
```


```{r fit tidy}
fit_type_tidied <- fit_type %>% tidy()
write_csv(fit_type_tidied, "brooklyn_fit_type_tidy.csv")
```

```{r model parameters}
report(fit_type %>% filter(nypd_pct_cd == 60, call_type == "CIP") %>% select(base_arima))
```

Extract fitted model 

```{r extract model data}
# first term is the model cols 3-8
# second term is the precinct/call type rows 1 - 23
fit_type[[4]][[1]][["data"]] # actual data with timestamps
#8,9
fit_type[[4]][[1]][["fit"]][["est"]] # fitted values with residuals

#[["par"]] = term, estimate, p-val
fit_type[[4]][[1]][["fit"]][["par"]]
```

```{r}
# first term is the model cols 3-8
# second term is the precinct/call type rows 1 - 23
fit_type[[8]][[1]][["data"]] # actual data with timestamps
#8,9
fit_type[[8]][[1]][["fit"]][["est"]] # fitted values with residuals

#[["par"]] = term, estimate, p-val
fit_type[[8]][[1]][["fit"]][["par"]]
```

```{r precinct diagnostics}
# fit_arima_exo %>% ggtime::gg_tsresiduals(lag = 21) + theme_minimal()
precinct_num = 61
type = "Non CIP"

fit_type %>% filter(nypd_pct_cd == precinct_num, call_type == type) %>% select(base_arima) %>% ggtime::gg_tsresiduals(lag = 21) +
  labs(title = paste0("Non CIP\nmint_arima Residuals - Precinct: ", precinct_num), x = "Date", y = "Call counts") + theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "lightgrey",
                                        color = "lightgrey",
                                        linewidth = 0),
        panel.grid.major = element_line(color = "grey", linewidth = 0.5),
        panel.grid.minor = element_blank(), # Removes minor grid lines
        strip.text = element_text(colour = "black"))

selected_precinct <- fit_type %>% filter(nypd_pct_cd == precinct_num, , call_type == type)

augment(selected_precinct) %>% features(.innov, ljung_box, lag = 14, dof = 6)
```


```{r ljung_box scores}
ljung_box_scores <- augment(fit_type) %>% features(.innov, ljung_box, lag = 14, dof = 6) %>% arrange(nypd_pct_cd, .model)
ljung_box_scores
```

Models with white noise residuals 

```{r}
ljung_box_scores%>% 
  filter(lb_pvalue >= 0.05) %>% 
  group_by(nypd_pct_cd, call_type) #%>% summarise(count = n())
```

### Call Type Forecasts

```{r forecast}
fc_type <- fit_type %>% forecast(daily_counts_type_ts %>% filter(incident_date > "2025-06-23"))
```


```{r fig.width = 8,fig.height = 8}

# Visualizing the forecasts of 1 precinct across all call types and agg

fc_type %>% filter(nypd_pct_cd == 61) %>% 
  autoplot(daily_counts_type_ts %>% 
             filter(incident_date >= "2025-06-15" & nypd_pct_cd == 61), 
           level = NULL) +
  facet_wrap(vars(call_type), 
             nrow = 3, 
             scales = "free_y", 
             labeller = labeller(
               call_type = c("<aggregated>" = "Aggregated",
               "CIP" = "CIP", 
              "Non CIP" =  "Non CIP")
               )) +
  labs(title = "Forecasts on 7 Day Test Set\nPrecinct 61", x = "Date", y = "Call counts") + 
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "lightgrey",
                                        color = "lightgray",
                                        linewidth = 0,
                                        linetype = "solid"),
        panel.grid.major = element_line(color = "grey", linewidth = 0.5),
        panel.grid.minor = element_blank(), # Removes minor grid lines
        strip.text = element_text(colour = "black"))
```


```{r view precinct 60 forecasts, fig.width = 12}
# Visualizing the forecasts of 1 precinct across selected call type

fc_type %>% filter(nypd_pct_cd == 61, call_type == "CIP") %>% 
  autoplot(daily_counts_type_ts %>% 
             filter(incident_date >= "2025-06-01" & nypd_pct_cd == 61), level = NULL) +
  labs(title = "Forecasts on 7 Day Test Set\nPrecinct 61", x = "Date", y = "Call counts") + 
  facet_wrap(vars(.model), scales = "free_y") + 
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "lightgrey"),
        panel.grid.major = element_line(color = "grey", linewidth = 0.5),
        panel.grid.minor = element_blank(), # Removes minor grid lines
        strip.text = element_text(colour = "black"))
```

```{r fig.width = 12,fig.height = 8}

# Visualizing the forecasts of 1 precinct across selected call type w/ PI

fc_type %>% filter(nypd_pct_cd == 61, call_type == "CIP") %>% 
  autoplot(daily_counts_type_ts %>% 
             filter(incident_date >= "2025-06-01" & nypd_pct_cd == 61)) +
  labs(title = "Forecasts on 7 Day Test Set\nPrecinct 61", x = "Date", y = "Call counts") + 
  facet_wrap(vars(call_type, .model), scales = "free_y") + 
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "lightgrey"),
        panel.grid.major = element_line(color = "grey", linewidth = 0.5),
        panel.grid.minor = element_blank(), # Removes minor grid lines
        strip.text = element_text(colour = "black"))
```

```{r view, fig.width = 16, fig.height = 10}
# Visualizing the forecasts of 3 precincts across both call types

fc_type %>% filter(nypd_pct_cd %in% c(61,67,71,94), call_type %in% c("CIP", "Non CIP")) %>% 
  autoplot(daily_counts_type_ts %>% 
             filter(incident_date >= "2025-6-01" & nypd_pct_cd %in% c(61,67,71,94)), color = "lightgrey") + 
  facet_wrap(vars(nypd_pct_cd, call_type), nrow = 4, ncol = 2, scales = "free_y") + 
  theme(plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        strip.background = element_rect(fill = "lightgrey"),
        panel.grid.major = element_line(color = "grey", linewidth = 0.5),
        panel.grid.minor = element_blank(), # Removes minor grid lines
        strip.text = element_text(colour = "black")) +
  labs(title = "Forecasts on 7 Day Test Set", x = "Date", y = "Call counts")
```

```{r save forecast}

write_csv(fc_type, "Output/brooklyn/brooklyn_forecast.csv")
```

### Prediction intervals 

Get prediction intervals for all models - all precincts - all call types

```{r prediction intervals}
fc_type_hilo <- fc_type %>% hilo() %>% select(`80%`, `95%`)
fc_type_hilo
write_csv(fc_type_hilo, "Output/brooklyn/brooklyn_fc_hilo.csv")
```

```{r}
accuracy_df <- accuracy(fc_type, data = daily_counts_type_ts)
write_csv(accuracy_df, "Output/brooklyn/brooklyn_fc_accuracy.csv")
accuracy_df %>% filter(nypd_pct_cd == 60)
```

## SAVE OUTPUTS

```{r}

# combine model metrics and residual evaluation
#fit_type %>% glance()

combined_acc_ljung <- fit_type %>% glance() %>% 
  arrange(nypd_pct_cd, .model) %>% 
  left_join(ljung_box_scores, 
            by = join_by(nypd_pct_cd == nypd_pct_cd,
                         call_type == call_type,
                         .model == .model))

combined_acc_ljung <- combined_acc_ljung %>%
  mutate(ar_roots_str = map_chr(ar_roots, ~paste(., collapse = "|")),
         ma_roots_str = map_chr(ma_roots, ~paste(., collapse = "|"))) %>%
  select(-c(ar_roots,ma_roots)) # Remove the original list column
```

```{r}
write_csv(combined_acc_ljung, "brooklyn_model_report.csv")
```



